Crunching Large Graphs with Commodity Processors

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
rebooting

%% move stuff to defs file
\newcommand{\todo}[1]{{\bfseries [[#1]]}}
%% To disable, just uncomment this line
 \renewcommand{\todo}[1]{\relax}

@@

\title{Can We Crunch the Social Graph on the Cheap?}

@@

%todo format the UW names horizontally
\authorinfo{Jacob Nelson, Brandon Myers, Andrew Hunter, Dan Grossman, Mark Oskin,
Carl Ebeling, Luis Ceze}{University of Washington}{\{nelson, bdmyers, ahh, djg, oskin, ebelin, luisceze\}@cs.washington.edu}
\authorinfo{Simon Kahan}{Pacific Northwest National Lab}{skahan@cs.washington.edu}
\authorinfo{Preston Briggs}{}{preston.briggs@gmail.com}

\maketitle
\begin{abstract}

 Techniques from graph analytics have found important applications in
  areas such as social networks and bioinformatics. These irregular
  problems present a challenge for parallel machines: obtaining
  performance is not straightforward. Multithreading works well for
  such problems, but the only large-scale implementation of the technique
  is the Cray XMT, an expensive, non-commodity machine.

  Our goal is to build a system with XMT-like properties, implemented
  with commodity components. This paper presents a runtime for latency
  tolerance, along with our plans for scaling up to multiple
  nodes. Our analysis shows the feasibility of our approach.
\end{abstract}

@@

Techniques from graph analytics have found important applications in
areas such as social networks and bioinformatics. \todo{more}

@@

The most interesting computational challenge comes from large,
low-diameter, power law graphs: this combination makes extracting
performance difficult. They do not fit in a single
commodity machine's memory. Their low diameter means that they are
difficult to lay out with locality: every vertex needs to be near every
other vertex. Their power law distribution means that a few vertices
generate much more work than the rest, leading to load
imbalances. Together, these properties lead to a
difficult conclusion: each edge traversal is likely to induce a very
expensive cache miss or remote access.

Multithreading is a technique that has been used successfully to
implement efficient computations for these graphs. The Cray XMT is an
example of such an approach: it solves the memory latency problem
through concurrency rather than caching. Each XMT processor supports
128 hardware contexts and 1024 outstanding memory operations, and is
able to switch contexts every cycle \todo{cite XMT description}. This ability comes at a cost: the
XMT is an expensive, non-commodity machine. \todo{don't we want to also say XMT has poor performance on x kind of apps?}

\begin{figure*}[htbp]
  \begin{center}
    \includegraphics[width=0.9\textwidth]{figures/system-overview.pdf}
	\end{center}
	\caption{Overview of our proposed system design. The base
          system is a cluster of commodity nodes and interconnect. We
          add a latency-tolerant runtime and a global memory manager
          to obtain efficient access to the global address space. Only
          the shaded components may require hardware design.}
	\label{fig:system-overview}
\end{figure*}

We want the best of both worlds. Our goal is to build a system that
has good performance on these low-locality graph codes but is
implemented with cheap commodity processors.

Figure~\ref{fig:system-overview} shows the structure of our
proposal. It is composed of multiple nodes built with commodity
processors, communicating over an Infiniband network. We add two
components: a runtime system, responsible for executing the user's
threads, and a global memory manager, responsible for facilitating
memory requests to the non-coherent global memory shared across the
nodes.  We are exploring a mix of hardware and software
implementations of the memory manager.

This paper explores the feasibility of our multi-node goal using a
single-node implementation of our lightweight threading runtime. We
use prefetch instructions together with lightweight threads to
tolerate the latency of DRAM access on the local node node. We use
pointer chasing benchmarks to verify our runtime's overhead is
acceptable. We also model the effects of pointer chasing on a remote
node.

The rest of the paper is organized as follows. We describe our
proposal from a programmer's viewpoint in section~\ref{sec:model}. We
describe the implementation of our latency-tolerant runtime and plans
for extending to multiple nodes in section~\ref{sec:approach}. We
evaluate our runtime in section~\ref{sec:evaluation}.  We present
related work in section~\ref{sec:related} and conclude in
section~\ref{sec:conclusion}.

@@

To evaluate the potential for our runtime to perform in the multi-node case, we simulated a network delay and saw whether the runtime could tolerate the larger latency.

Since the network interface is accessed over QPI, we estimated QPI bandwidth by measuring the QPI communication between the two chips of our test machine. To do this, we allocated the lists in the second chip's memory and ran the traversals on the first chip's cores. We found that use of the QPI link limits us to 175 \mrps.

@@

To simulate the performance of pointer chasing in a
multi-node system, we modified the scheduler of our coroutine library
to include a delay before a coroutine can be rescheduled, imitating
the network transit delay even though we are referencing local
memory. Recently published measurements of Infiniband latencies
\todo{cite} suggest we might expect 1.1 $\mu$s source-to-destination
latencies between two computers connected back-to-back; we assume a
switch transit delay of 400 ns \todo{how many hops? 1or3?} in each direction and thus estimate a 3
$\mu$s round-trip delay.

%simulated delay plot
\begin{figure}[h]
  \begin{center}
    \includegraphics[width=0.47\textwidth]{figures/multi-green-delay7800-edited.pdf}
  \end{center}
  \vspace{3pt}
  \caption{Pointer chasing with simulated network delay. Each core ran up to 167 coroutines, and each coroutine made 1 to 4 concurrent references.}
  \label{fig:network-delay}
\end{figure}

The results of the experiment are shown in
Figure~\ref{fig:network-delay}. We are still able to reach a
near-maximum rate of 273 \mrps\footnote{The throughput keeps increasing up to around 850 total concurrent references offered, far more than the 36 we know the chip to support. This is because we have modeled the network latency by forcing coroutines to wait the extra time before using the data, so the physical miss buffers on our machine are not tied up any longer than usual.}, but at least 2
concurrent references per coroutine are required. Running with a single
reference per coroutine achieves 236 \mrps.

@@

%\todo{Begin eval insert}
%\input{evaluation}
%\todo{End eval insert}

\section{Related work}
\label{sec:related}
The concept of using independent parallel contexts to tolerate latency
is not novel.  The inspiration for our project is of course the Cray
XMT \cite{feo-xmt}, which implemented a hardware version of our
programming model (in what would today be considered exceptionally
slow silicon.)  The XMT can be seen as simultaneous multithreading
\cite{tullsen-smt} writ large; the technique and motivation is
essentially the same, but the XMT allows hundreds of contexts (modern
Intel SMT implementations are generally limited to two.) The Alewife
system \cite{agarwal-alewife} also switches contexts (in hardware) to
hide latency on remote-node accesses.  Cyclops \cite{almasi-cyclops}
allows hundreds of independently executing contexts to share FPUs,
caches, and paths to memory, but does not context switch or share
basic execution hardware.

Modern programmable GPUs, such as nVidia's CUDA programming model, can also be interpreted in using an XMT-like
model under the hood;
% Do I need to include a citation here?
while they generate large numbers of available execution contexts via
a (semi-)explicit SIMD model, those SIMD instructions keep functional
units satisfied by overlapping execution with the loads required for
other contexts.

Similar techniques have also been used in software.  Mowry proposed a system \cite{mowry-scm} of \emph{software-controlled
  multithreading}.  As opposed to our system of explicitly marked
remote references, SCM assumes a high hit rate and traps to an
interrupt routine which performed a context switch upon a cache miss.  The 2000-era processors had
short enough latencies that using more than two contexts generally
produced slowdown.

Partitioned global address space models, such as that used in Unified Parallel C,
% definitely need a citation here
 share our approach of exposing all data (even remote data)
 transparently to any threads and have largely solved the problems of
 how to efficiently translate those remote reads into message
 passing.  However, their programming models emphasize locality more
 than our work does; while a program can examine any vertex in a UPC
 program at any time, it probably shouldn't (whereas our kernels rely on our
 latency hiding technique to make common remote references fast.

The value of the information in extremely large and sparse graphs is
rapidly growing.  The GraphCT toolkit \cite{ediger-graphct} used an XMT to perform
sophisticated analysis on data from Twitter, identifying important
actors and their networks of influence, and it can scale to the full
size of graphs such as Facebook's friend network.

>>>>

%% move stuff to defs file
\newcommand{\todo}[1]{{\bfseries [[#1]]}}
%% To disable, just uncomment this line
% \renewcommand{\todo}[1]{\relax}

@@

%\textheight 9in
%\textwidth 6.5in
%\columnsep 1pc
\renewcommand{\baselinestretch}{0.9}

%\title{Can We Crunch the Social Graph on the Cheap?}
\title{Crunching Large Graphs with Commodity Processors}

@@

%todo format the UW names horizontally
\authorinfo{Jacob Nelson, Brandon Myers, A.H. Hunter, Preston Briggs,
  Dan Grossman, Mark Oskin, Carl Ebeling, Simon Kahan, Luis
  Ceze}{University of Washington}{\{nelson, bdmyers, ahh, preston,
  djg, oskin, ebeling, skahan, luisceze\}@cs.washington.edu}

\maketitle

\begin{abstract}
% Techniques from graph analytics have found important applications in
% areas such as social networks and bioinformatics. These irregular
% problems present a challenge for parallel machines: obtaining
% performance is not straightforward. Multithreading works well for
% such problems, but the only large-scale implementation of the technique
% is the Cray XMT, an expensive, non-commodity machine.

% Our goal is to build a system with XMT-like properties, implemented
% with commodity components. This paper presents a runtime for latency
% tolerance, along with our plans for scaling up to multiple
% nodes. Our analysis shows the feasibility of our approach.

  Crunching large graphs is the basis of many emerging applications,
  such as social network analysis and bioinformatics. Graph analytics
  algorithms exhibit little locality and therefore present significant
  performance challenges. Hardware multithreading systems (e.g, Cray
  XMT) showed that with enough concurrency, we can tolerate long
  latencies. Unfortunately, this solution is not available with
  commodity parts.

  Our goal is to develop a latency-tolerant system built out of
  commodity parts and mostly in software. The proposed system includes
  a runtime that supports a large number of lightweight contexts,
  full-empty synchronization and a memory manager that provides a
  high-latency but high-bandwidth global shared memory. This paper
  lays out the vision for our system, and justifies its feasibility
  with a performance analysis of the runtime for latency tolerance.

% Our analysis shows the feasibility of our
%   approach.

% To that end we developed a new runtime system that supports
%   a large number of lightweight contexts, together with full-empty
%   synchronization primitives and a large global address space.

% To that end, we propose a
%   new runtime system that is able to sustain a large number of active
%   threads

%  XMT-like properties, implemented
%   with commodity components. This paper presents a runtime for latency
%   tolerance, along with our plans for scaling up to multiple
%   nodes. Our analysis shows the feasibility of our approach.

%   Large, sparse, irregular graphs map poorly to the memory systems of
%   commodity hardware. The scale of emerging real world datasets
%   presents a challenge for straightforward algorithms which will run
%   inefficiently due to many high-latency accesses. Pervasive
%   parallelism can tolerate that latency, but the only large-scale
%   implementation of the technique is the Cray XMT, an expensive,
%   non-commodity machine.

%   If we can adapt the XMT's programming model to commodity hardware,
%   we can dramatically improve performance without resorting to
%   expensive custom systems. This paper presents a runtime for latency
%   tolerance through the use of parallel lightweight contexts, along
%   with our plans for scaling up to multiple nodes. Our analysis shows
%   the feasibility of our approach.
\end{abstract}

@@

Many important emerging applications such as social network analysis,
bioinformatics and analysing vast sensor data from actual networks
(e.g, water, electrical grid) are based on crunching very large
graphs. Unfortunately, the computational cost of these applications
gets quickly out of hand. For example, in January 2011, Facebook had
500 million active users with an average of 130 friends each
\cite{Facebook:2011p91}. Its friend suggestion algorithm in 2010 ran on
  a 40-node cluster with 72GB memory per node, and still, new
  suggestions could only be computed every two days
  \cite{Backstrom:2010p90}. Speeding up these applications at a low
  cost would have a large impact in how we deal and react to the fast
  growth of the data-set and would make the results even more
  valuable.

% Techniques from graph analytics have found important applications in
% areas such as social networks and bioinformatics. \todo{more!}

@@

In this context, the most interesting computational challenge comes
from large, low-diameter, power law graphs: this combination makes
extracting performance difficult. They do not fit in a single
commodity machine's memory. Their low diameter means that they are
difficult to lay out with locality: every vertex needs to be near
every other vertex. Their power law distribution means that a few
vertices generate much more work than the rest, leading to load
imbalances. Together, these properties lead to a difficult
conclusion: each edge traversal is likely to lead to expensive
inter-node communication. %cache miss or remote access.

Multithreading is a technique that has been used successfully to
implement efficient computations for these graphs. The Cray XMT is an
example of such an approach: it solves the memory latency problem (or
communication problem) through concurrency rather than caching. Each
XMT processor supports 128 hardware contexts and 1024 outstanding
memory operations, and is able to switch contexts every cycle.
% \todo{cite XMT description}.  %% LUIS: there is enough information here.
This ability comes at a cost: the XMT is an expensive, non-commodity
machine. Moreover, arguably a lot of XMT's advantages for these type
of applications come at the cost of performance in general-purpose
applications.

% \todo{don't we want to also say XMT has poor performance on x
%   kind of apps?}

\begin{figure*}[htbp]
  \begin{center}
    \vspace{-0.25in}
    \includegraphics[width=0.7\textwidth]{figures/system-overview.pdf}
    \vspace{-0.2in}
	\end{center}
	\caption{Overview of our proposed system design. The base
          system is a cluster of commodity nodes and interconnect. We
          add a latency-tolerant runtime and a global memory manager
          to obtain efficient access to the global address space. Only
          the shaded components may require hardware design.}
	\label{fig:system-overview}
\end{figure*}

%We want the best of both worlds. 
We believe we can build a system mostly based on commodity parts that
can attain XMT-like performance with a familiar, also XMT-like,
programming model, but at a fraction of the cost. Therefore, our goal
is to build a system that has good performance on low-locality graph
codes but is implemented with cheap commodity processors and simple
additional support in an FPGA. This approach also has the added
benefit of not sacrificing general-purpose performance.

Figure~\ref{fig:system-overview} shows the structure of our
proposal. It is composed of multiple nodes built with commodity
processors, communicating over an Infiniband network. We add two
components: a runtime system, responsible for executing the user's
threads, and a global memory manager, responsible for facilitating
memory requests to the non-coherent global memory shared across the
nodes.  We are exploring a mix of hardware (FPGA) and software
implementations of the memory manager.

This paper describes the vision of our system and explores the
feasibility of our ideas using a single-node implementation of a
lightweight threading runtime. We use prefetch instructions together
with lightweight threads to tolerate the latency of DRAM access on the
local node node. We use adversarial pointer chasing benchmarks to
verify our runtime's overhead is acceptable. We also model the effects
of pointer chasing on a remote node by artificially increasing
latencys to the order of several micro-seconds.

The rest of the paper is organized as follows. We quickly describe our
programming model in section~\ref{sec:model}. We describe the
implementation of our latency-tolerant runtime and plans for extending
to multiple nodes in section~\ref{sec:approach}. We evaluate our
runtime in section~\ref{sec:evaluation}.  We present related work in
section~\ref{sec:related} and conclude in
section~\ref{sec:conclusion}.

@@

To evaluate the potential for our runtime to perform in the multi-node case, we simulated a network delay and saw whether the runtime could tolerate the larger latency. Since the network interface is accessed over QPI, estimated QPI bandwidth by measuring the QPI communication between the two chips of our test machine.

@@

For this experiment, we allocated the lists in the
second chip's memory and ran the traversals on the first chip's cores. We found that use of the QPI link limits us to 175 \mrps.

To simulate the performance of pointer chasing in a
multi-node system, we modified the scheduler of our coroutine library
to include a delay before a coroutine can be rescheduled, imitating
the network transit delay even though we are referencing local
memory. Recently published measurements of Infiniband latencies
\todo{cite} suggest we might expect 1.1 $\mu$s source-to-destination
latencies between two computers connected back-to-back; we assume a
switch transit delay of 400 ns \todo{how many hops? 1or3?} in each direction and thus estimate a 3
$\mu$s round-trip delay. In this experiment, once a coroutine has
yielded, we use the core's timestamp counter to ensure it does not get
rescheduled for 3 $\mu$s.

%simulated delay plot
\begin{figure}[h]
  \begin{center}
    \includegraphics[width=0.47\textwidth]{figures/multi-green-delay7800-edited.pdf}
  \end{center}
  \caption{Pointer chasing with simulated network delay.}
  \label{fig:network-delay}
\end{figure}

We ran these experiments with two sources of memory concurrency: each
core ran up to 167 coroutines, and each coroutine made between 1 and 4
concurrent references. The results are shown in
Figure~\ref{fig:network-delay}. We are still able to reach a
near-maximum rate of 273 \mrps, but at least 2
concurrent references per coroutine are required. Running with a single
reference per coroutine achieves 236 \mrps. Note that the throughput keeps increasing up to around 850 total concurrent references offered, far more than the 36 we know the chip to support. This is because we have modeled the network latency by forcing coroutines to wait the extra time before using the data, so the physical miss buffers on our machine are not tied up any longer than usual.

@@

\section{Related work}
\label{sec:related}

Much work exists on using multithreading to tolerate latency. Hardware
implementations include the Tera MTA~\cite{tera} and Cray
XMT~\cite{feo-xmt}, Simultaneous Multithreading \cite{tullsen-smt},
MIT Alewife \cite{agarwal-alewife}, Cyclops \cite{almasi-cyclops}, and
even GPUs \cite{gpus}. Software-only approaches exist as well; the
Software Controlled Multithreading project \cite{mowry-scm}, QThreads
\cite{qthreads}, MAESTRO \cite{maestro}, and Charm \cite{charm} all
use variants of this idea.
%In contrast to all of these, our goal is to tolerate
%latencies across a multi-node system using mostly software.

There has been much work on user level threading; the First-Class User
Level Threading project \cite{ult} is one example. Our goal is
slightly different; we want many contexts to tolerate memory latency.

Our goal of presenting a global view of distributed memory to the
programmer is shared by the PGAS community, and is used in languages
like Chapel \cite{chapel}, X10 \cite{X10}, and UPC \cite{upc}. They
obtain performance by minimizing references to remote nodes, whereas we
optimize for this case.

The idea of processing large graphs on distributed machines has been
exploredby projects such as Pregel \cite{pregel}, the Parallel Boost
Graph Library \cite{parallelbgl}, and CGMLib \cite{cgmlib}. Our goal
is to develop infrastructure to aid implementation of similar
libraries.
